Novel lattice Boltzmann method for simulation of strongly shear thinning viscoelastic fluids

### 2024-01-16 NovelLBM.tex
introduction

	\section{Introduction}
	\todo{do}

@@

	With the validity and accuracy of our scheme firmly established, we can demonstrate its results for some systems that cannot be solved analytically. In the following we will show the flow profiles for a conical nozzle, a cell in shear flow, and a chip for RT-DC measurements\todo{cite? spell our RT-DC?}. For all of these the methyl cellulose parameters found for a concentration of \SI{0.49}{\percent} have been used.

@@

	Real-time deformability cytometry (RT-DC)\todo{cite} is a novel approach for single cell characterization. The cells are forced through a constriction. The observed deformation can be used to infer the mechanical properties of the cell. For this simulation we use periodic boundary conditions and a pressure gradient\todo{specify}. The resulting flow can be seen in figure \ref{fig:RT-DC_v}. The component typically associated with viscosity of the stress tensor can be see in figure \ref{fig:RT-DC_tau}. There are three qualitative differences compared to a purely viscous fluid. Directly after the inlet is a region of low stress. Due to the advection of the stress, the fluid here despite being inside the constriction has not reacted to it yet. This region therefore could produce unexpected cell behavior if one passes through it. Similarly directly after the constriction, the stress has not yet lowered. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find an extremely low and an extremely high viscosity respectively. Depending on the exact parameters \glqq viscosities\grqq{ } can be reached that cannot be obtained through rheological measurements with the fluid. Meaning these viscosities are not valid. This is even clearer along the walls directly after the outlet. Here the viscosites obtained in this manner become negative. This shows the breakdown of the viscosity interpretation in viscoelastic cases and signifies the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by shear thinning models and should be considered despite the high computational cost.

>>>>

	\section{Introduction}
	For a solution to be well suited for usage as a bioink during bioprinting its rheology has to fulfill certain requirements. During the printing process it should flow easily through the nozzle, meaning the viscosity should be low. This is not only beneficial for ease of use, but is also crucial for survivability of the cells, as the stress they experience is dependent on the viscosity. However, after the printing process, it is desired, that the printed construct is stable. This can be temporarily archived by very high viscosities. Therefore, one approach to solve this is to use shear thinning fluids. Shear thinning fluids can easily be produced by dissolving polymers in water. This however, has the side effect of viscoelasticity. Due to the polymers the solution also exhibits elastic behavior. The influence of which on the printing process is not well understood yet. For realistic bioinks, the ratio of their zero shear viscosity and their infinite shear viscosity is large as is expected based on the desired attributes mentioned above. For such large viscosity ratios, simulating viscoelastic materials with Lattice Boltzmann methods was to our knowledge not yet possible.

	Such simulations would however be very beneficial. They allow the \glqq measurement\grqq{ }of parameters that are otherwise hard or even impossible to access. They allow the prediction of the outcome of experiments without performing them, which is especially helpful if the experiment cannot be easily performed. An example would be the design of new and improved printing nozzles. With simulations, the geometry can be optimized to a degree, that would be near impossible using iterative experiments. The large amounts of data simulations can provide allow better understanding of the fundamental processes, that cause observed behavior. \todo{"List some simulations for bionprinting, bioinks that have been done (restrict to only viscoelastic sims)"}

	Simulations using Lattice Boltzmann methods (LBM) are popular, as it has many advantages. It is very extensible and thus can deal with complex geometries, particles (e.g. cells) suspended in the fluid and many different interactions within and between those. Most importantly it is and easily parallelizable algorithm and thus can be run on GPUs or similar accelerators. This allows even large domains to be simulated in short amount of time with commodity GPUs and no need for any supercomputer. It is also easily possible to extend LBM to deal with viscoelastic fluids. This has been done before by multiple groups. There are many publications discussing such approaches in two dimensions\cite{ispolatov2002,wang2019,su2013,giraud1998,xie2018,dzanic2022,frank2006,zou2014,lee2017,dzanic2022_2,sedaghat2021,frantziskonis2011,qin2023,yoshino2008,papenkort2015,phillips2011,onishi2005}. Simulations in three dimensions have a higher computational cost and are not as abundant, but have also been performed by many\cite{malaspinas2010,gupta2015,lallemand2003,ma2020,onishi2006,wang2020,dellar2014,su2013_2,wang2021,kuron2021,gupta2016}. Among these many are concerned about stability. Some include diffusivity of the polymer stress in their model to increase stability\cite{malaspinas2010,ma2020,wang2020,wang2021}. Other remedies include special decompositions of the stress tensor\cite{dzanic2022} or optimization of the solvent contribution\cite{gupta2015}. However, none of these reach the relevant viscosity ratios for bioinks.

	Here we show our novel simulation approach for bioinks. It is specifically tailored to support arbitrary viscosity ratios and thus is able to simulate common bioinks or any other viscoelastic material with similar properties while providing high accuracy and stability. This is archived by subtracting a viscous stress from the polymer stress calculated from the PTT model. This offset is corrected by increasing the viscosity of the base LBM algorithm accordingly.

@@

	With the validity and accuracy of our scheme firmly established, we can demonstrate its results for some systems that cannot be solved analytically. In the following we will show the flow profiles for a conical nozzle, a cell in shear flow, and a chip for RT-DC measurements\todo{cite? spell out RT-DC?}. For all of these the methyl cellulose parameters found for a concentration of \SI{0.49}{\percent} have been used.

@@

	Real-time deformability cytometry (RT-DC)\todo{cite} is a novel approach for single cell characterization. The cells are forced through a constriction. The observed deformation can be used to infer the mechanical properties of the cell. For this simulation we use periodic boundary conditions and a pressure gradient\todo{specify}. The resulting flow can be seen in figure \ref{fig:RT-DC_v}. The component typically associated with viscosity of the stress tensor can be see in figure \ref{fig:RT-DC_tau}. There are three qualitative differences compared to a purely viscous fluid. Directly after the inlet is a region of low stress. Due to the advection of the stress, the fluid here despite being inside the constriction has not reacted to it yet. This region therefore could produce unexpected cell behavior if one passes through it. Similarly directly after the constriction, the stress has not yet lowered. If one would divide the stress in these two regions by the strain rate in an attempt to obtain a viscosity one would find an extremely low and an extremely high viscosity respectively. Depending on the exact parameters \glqq viscosities\grqq{ }can be reached that cannot be obtained through rheological measurements with the fluid. Meaning these viscosities are not valid. This is even clearer along the walls directly after the outlet. Here the viscosites obtained in this manner become negative. This shows the breakdown of the viscosity interpretation in viscoelastic cases and signifies the regions of interesting behavior. This also shows why viscoelastic simulations cannot always be replaced by shear thinning models and should be considered despite the high computational cost.
